\section{Conclusion}
\label{sec_concl}

This paper described the TPTP TFF1 format, an extension of the monomorphic TFF0 format
with rank-1 polymorphism. The new format complements the existing TPTP
offerings. %, taking over where TFF0 left off.
For reasoning tools that already
support polymorphism, TFF1 is a portable alternative to the existing ad hoc
syntaxes. But more importantly, the format is a vehicle to foster native
polymorphism support in automatic theorem provers. The time is ripe: After many
years of untyped reasoning, we have recently witnessed the rise of
interpreted arithmetic embedded in monomorphic type systems. TFF1
lifts the most obvious restrictions of such type systems.

TFF1 is part of TPTP World. The TPTP library already contains
nearly a thousand TFF1 problems, and although the format is in its
infancy, it is supported by several applications, including the SMT solver
Alt-Ergo (via Why3).
Work has commenced in Saarbr\"ucken to add polymorphism to the
superposition prover SPASS \cite{wand-weidenbach-2012}.
Given that many applications require polymorphism, other reasoning tools are
likely to follow suit.
The annual CADE Automated System Competition (CASC) will certainly have a
role to play driving adoption of the format. But regardless of progress in
prover technology, equipped with a concrete syntax and suitable middleware,
users can already turn their favorite automatic theorem prover into a
fairly efficient polymorphic prover.

Rank-1 polymorphism is, of course, no panacea. More advanced features, such as
type classes, higher ranks, and dependent types, are not catered for (although type
classes can be comfortably encoded in TFF1). These are expected to be
part of a future TFF2, with the proviso that there be sufficient interest from
users and implementers.

\def\ackname{Acknowledgment}
\paragraph{\textbf{\upshape\ackname.}}
%
The present specification is largely the result of consensus among
participants of the {\tt polymorphic-tptp-tff} mailing list, especially
Fran\c{c}ois Bobot, Chad Brown, Florian Rabe, Philipp R\"ummer, Stephan Schulz,
Geoff Sutcliffe, and Josef Urban.
We are grateful to Geoff Sutcliffe, TPTP Master of Ceremonies, for giving TFF1
his benediction and adapting the TPTP BNF and other infrastructure.
He, Mark Summerfield, and several anonymous reviewers suggested many textual
improvements to this paper. We also thank Viktor Kuncak, Tobias Nipkow, and
Nicholas Smallbone for their support and ideas.
%
The first author's research was supported by the Deutsche
Forschungs\-gemein\-schaft project \relax{Hardening the Hammer} (grant
Ni\,491\slash 14-1).
%
%The authors are listed in alphabetical order regardless of individual
%contributions or seniority.
